Graph Topological Aspects of Granger Causal Network Learning\\ \large Supplementary Material

### 2019-07-23 paper_theory/proofs.tex
Substantial updates to the proof of the confounding theorem.

\begin{definition}[Strongly Connected Components]
  \label{def:scc}
  In a graph $\gcg$, the subset $S \subseteq [n]$ is \textit{strongly
    connected} if $\forall i, j \in S$, $i \in \anc{j}$ and
  $j \in \anc{i}$.  We will denote by $S(j)$ (which may be the
  singleton $\{j\}$) the largest strongly connected component (SCC)
  containing $j$.  We will denote $x_{S(j)}(t)$ to be the vector of
  processes

@@

  whose indices are sorted in ascending order.  Similarly, the
  sub-filter of $\B(z)$ acting on $x_{S(j)}(t)$ will be denoted
  $\B_{S(j)}(z)$.
\end{definition}

@@

  \begin{equation}
    \label{eqn:scc_parent_expansion_inverted}
    x_{S(j)}(t) = (I - \B_{S(j)}(z))^{-1} \big(v_{S(j)}(t) + \sum_{\substack{s \in S(j) \\ k \in \pa{s} \cap S(j)^\c}} B_{sk}(z)x_k(t)e_s^{S(j)}\big),
  \end{equation}

@@

  \[
    x(t) = \B(z)x(t) + v(t).
  \]

  \hl{TODO: show stability}
  % For the stability of $\B_{S(j)}(z)$ we must show that
  % $\forall z: |z^{-1}| \le 1$ we have $$||\B_{S(j)}(z)||_2 < 1.$$
\end{proof}

@@

  \begin{align*}
    &\inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{(i)}}}\\
    &= \E|F_{r + 1}(z)x_j(t) - \linE{F_{r + 1}(z)x_j(t)}{\H_{t - 1}^{(i)}}|^2\\
    &+ \inner{F_{r + 1}(z)x_j(t) - \linE{F_{r + 1}(z)x_j(t)}{\H_{t - 1}^{(i)}}}{\sum_{\ell = 0}^r F_\ell(z) \wtalpha{a_\ell}{a_{\ell + 1}}},
  \end{align*}

  which by the Cauchy-Schwarz inequality is $0$ if and only if

@@

\begin{lemma}
  \label{lem:time_lag_cancellation}
  Suppose $v(t)$ is a scalar process with unit variance and zero
  autocorrelation and let $\A(z), \B(z)$ be nonzero and strictly
  causal (i.e. $\tau_0(\A) \ge 1$) linear filters.  Then,

@@

  \begin{align}
    \inner{\A(z)v(t)}{\B(z)v(t)} &= \sum_{\tau = 1}^\infty \sum_{s = 1}^\infty a(\tau)b(s)\E[v(t - s)v(t - \tau)]\\
    &= \sum_{\tau = \text{max}(\tau_0(\A), \tau_0(\B))}^{\text{min}(\tau_\infty(\A), \tau_\infty(\B))} a(\tau) b(\tau)\\
  \end{align}

@@

\begin{proposition}
  \label{prop:persistence_converse}
  Suppose $\gcg$ is a strongly causal DAG and that $x(t)$ is
  persistent, then if there exists a $k$ which confounds $(i, j)$ we
  have $i \pwgc j$ and $j \pwgc i$.
\end{proposition}
\begin{proof}
  We will show that $j \pwgc i$, the other being symmetric.  First
  note that by the contrapositive of proposition
  \ref{prop:sc_graph_common_anc} we cannot have $i \in \anc{j}$ or
  $j \in \anc{i}$ (else $k$ would not be a confounder) and therefore
  \textit{every} $u \in \anc{i}\cap\anc{j}$ will be a confounder.

  It is sufficient to show that $\exists \psi \in \H_{t - 1}^{(j)}$
  such that

  \begin{equation*}
    \inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{(i)}}} \ne 0.
  \end{equation*}

  To this end, let $F(z)$ be an arbitrary but strictly causal linear
  filter.  We apply equation \eqref{eqn:ancestor_expansion} to
  $x_i(t)$ and $\psi \defeq F(z)x_j(t)$:

  \begin{align*}
    &\inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{(i)}}}\\
    &\overset{(a)}{=} \inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{\A_{ii}(z)v_i(t) + \sum_{k \in \anc{i}}\A_{ik}(z)v_k(t)}\\
    &\overset{(b)}{=} \inner{\sum_{k \in \anc{j}}\big(F(z)\A_{jk}(z)v_k(t) \\&- \linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}}\big)}{\A_{ii}(z)v_i(t) + \sum_{k \in \anc{i}}\A_{ik}(z)v_k(t)}\\
    &\overset{(c)}{=} \sum_{k \in \anc{i}\cap\anc{j}}\Big(\inner{F(z)\A_{jk}(z)v_k(t)}{\A_{ik}(z)v_k(t)} \\&- \inner{\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}}}{\A_{ii}v_i(t)}\\
    &- \sum_{\ell \in \anc{i}}\inner{\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}}}{\A_{i\ell}(z)v_\ell(t)}\Big)
  \end{align*}

  where in $(a)$ we have removed the $\linE{x_i(t)}{\H_{t - 1}^{(i)}}$
  term via the orthogonality principle, in $(b)$ there is no
  $F(z)\A_{jj}(z)v_j(t)$ term since due to $j \not\in \anc{i}$ it is
  orthogonal to $\H_t^{(i)}$.  Finally, $(c)$ follows by applying
  orthogonality properties of $v(t)$, as well as the fact that
  $\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} = 0$ for
  $k \not \in \anc{i}$.  Note that
  $\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} \in \H_{t - 1}^{(i)}$
  and therefore there is in general no cancellation in the final term
  above for $\ell \in \anc{i}$.

  \todo{This is clearly not immediately evident}

  This is $0$ for every $F$ if and only if for all $F$ and
  $\forall k \in \anc{i} \cap \anc{j}$ we have
  $$\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} = 0$$ and
  $$\inner{F(z)\A_{jk}(z)v_k(t)}{\A_{ik}(z)v_k(t)} = 0,$$ which by
  Corollary \ref{cor:time_lag_cancellation} occurs if and only if
  $\tau_0(\A_{jk}) \ge \tau_\infty(\A_{ik})$, which is impossible since by persistence
  $\tau_0(\A_{jk}) < \infty$ and  $\tau_\infty(\A_{ik}) = \infty$.
\end{proof}

>>>>

\begin{definition}[Strongly Connected Components]
  \label{def:scc}
  In a graph $\gcg$, the \textit{ordered} (by the natural ordering on
  $\N$) subset $S \subseteq [n]$ is \textit{strongly connected} if
  $\forall i, j \in S$, $i \in \anc{j}$ and $j \in \anc{i}$.  We will
  denote by $S(j)$ (which may be the singleton $(j)$) the largest
  strongly connected component (SCC) containing $j$.  We will denote
  $x_{S(j)}(t)$ to be the vector of processes

@@

  whose indices are given the same (natural) ordering as $S(j)$.
  Similarly, the sub-filter of $\B(z)$ acting on $x_{S(j)}(t)$ will be
  denoted $\B_{S(j)}(z)$.
\end{definition}

@@

  \begin{equation}
    \label{eqn:scc_parent_expansion_inverted}
    x_{S(j)}(t) = (I - \B_{S(j)}(z))^{-1} \big(v_{S(j)}(t) + \sum_{\substack{s \in S(j) \\ k \in \pa{s} \cap S(j)^\c}} B_{sk}(z)x_k(t)e_s^{S(j)}\big).
  \end{equation}

@@

  \[
    x(t) = \B(z)x(t) + v(t),
  \]

  which, when rearranged appropriately, can be written as

  \[
    \left[
      \begin{array}{c}
        x_{S(j)}(t)\\
        x_{S(j)^c}(t)
      \end{array}
    \right] =
    \left[
      \begin{array}{cc}
        \B_{S(j)}(z)& \B_{S(j), S(j)^\c}(z)\\
        \B_{S(j)^\c, S(j)}(z)& \B_{S(j)^\c}(z)
      \end{array}
    \right]
        \left[
      \begin{array}{c}
        x_{S(j)}(t)\\
        x_{S(j)^c}(t)
      \end{array}
    \right] +
    \left[
      \begin{array}{c}
        v_{S(j)}(t)\\
        v_{S(j)^c}(t)
      \end{array}
    \right].
  \]

  Theorem \ref{thm:granger_causality_equivalences} is invoked in order
  to restrict the summation to $k \in \pa{s}$ (since other elements
  are $0$).

  Now, we can partition $\gcg$ into it's maximal SCCs
  $S_1, \ldots, S_N$, (one of which is $S(j)$) and then consider the
  DAG formed on $N$ nodes with edges $I \rightarrow J$ included on the
  condition that
  $\exists j \in S_J, i \in S_I \text{ s.t. } i \in \anc{j}$.  By
  topologically sorting this DAG, we obtain an ordering $\sigma$ of
  $[n]$ such that $\B_\sigma(z)$ is block upper triangluar, with one
  of it's diagonal blocks consisting of the (possibly reordered)
  matrix $\B_{S(j)}(z)$.  So we have

  \begin{align*}
    \forall\ |z^{-1}| \le 1: \det \B(z) = \prod_{i = 1}^N \det \B_{S_i}(z) &\ne 0\\
    \implies \forall\ |z^{-1}| \le 1: \det \B_{S(j)}(z) &\ne 0,
  \end{align*}

  and therefore $\B_{S(j)}(z)$ is stable, invertible, and Equation
  \eqref{eqn:scc_inversion} holds.
\end{proof}

@@

  \begin{align*}
    &\inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{(i)}}}\\
    &= \E|F_{r + 1}(z)x_j(t) - \linE{F_{r + 1}(z)x_j(t)}{\H_{t - 1}^{(i)}}|^2\\
    &+ \inner{F_{r + 1}(z)x_j(t) - \linE{F_{r + 1}(z)x_j(t)}{\H_{t - 1}^{(i)}}}{\sum_{\ell = 0}^r F_\ell(z) \wtalpha{a_\ell}{a_{\ell + 1}}}.
  \end{align*}

  Now since $F_{r + 1}(z) \ne 0$ by Theorem
  \ref{thm:granger_causality_equivalences}, and
  $F_{r + 1}(z)x_j(t) \not\in \H_{t - 1}^i$, we have by the
  Cauchy-Schwarz inequality that this expression is equal to $0$ if
  and only if

@@

\begin{example}
  Consider a process $x(t)$ generated by the $\VAR(1)$
  model\footnote{Recall that any $\VAR(p)$ model with $p < \infty$ can
    be written as a $\VAR(1)$ model, so we lose little generality in
    considering this case.}  having $\B(z) = Bz^{-1}$.  If $B$ is
  diagonalizable, and has at least $2$ distinct eigenvalues, then
  $x(t)$ is persistent.

  Pick any $i \in [n], j \in \anc{i} \setminus \{i\}$.  Then the
  stability of $B$ allows us to write

  \begin{equation*}
    \A(z) = \sum_{k = 0}^\infty B^k z^{-k},
  \end{equation*}

  whereby we see that $\exists k > 0$ such that $[B^k]_{ij} \ne 0$
  (since $j \in \anc{i}$).  Then consider

  \begin{equation*}
    \begin{aligned}
      [B^{rk}]_{ij} &= e_i^\T B^{rk} e_j \\
      &\overset{(a)}{=} \big((P^\T e_i)^\T J^{rk} P^{-1}e_j\big)\\
      &= \tr [(P^\T e_i)^\T J^{rk} P^{-1}e_j]\\
      &\overset{(b)}{=} \tr [(J^{rk}) (v u^\T)],
    \end{aligned}
  \end{equation*}

  where $(a)$ utilizes the Jordan Normal Form of $B$, and $(b)$
  denotes $u = P^\T e_i$ and $v = P^{-1}e_j$.  In order for
  $\tau_\infty(\A_{ij}) < \infty$, there must be some $N > 1$ such
  that $\forall r \ge N$, the above term is $0$.  This may be the case
  for instance if $B$ is a nilpotent matrix.

  Using the supposition that $B$ is diagonalizable (i.e. $J$ is a
  diagonal matrix) with at least $2$ distinct eigenvalues (in this
  case $B$ is \textit{not} nilpotent), we can then rewrite the above
  as

  \begin{equation*}
    f(r) \defeq \tr [(J^{rk}) (v u^\T)] = \sum_{\nu = 1}^n \lambda_\nu ^{rk} v_\nu u_\nu \defeq \sum_{\nu = 1}^n \lambda_\nu^{rk} \beta_\nu
  \end{equation*}

  where $\lambda_\nu$ denotes the eigenvalues of $B$ and
  $\beta_\nu = u_\nu v_\nu$.  Note that $f(0) = 0$ since $i \ne j$ and
  $u$ is a row of $P$ and $v$ is a column of $P^{-1}$.  Moreover,
  $f(1) \ne 0$ by hypothesis.  But, in order for
  $f(r) = 0\ \forall r \ge N$, it would need to be the case that

  \begin{equation*}
    \Dg(\bm{\lambda})^r \bm{\lambda} = Vz
  \end{equation*}

  had a solution in $z$ for every $r \ge N$, where $V$ is an
  $n \times n - 1$ full-rank matrix whose columns span the nullspace of $\beta$,
  and $\bm{\lambda} = (\lambda_1, \ldots, \lambda_n)$. That is,
  iterates of $\Dg(\bm{\lambda})$ applied to $\bm{\lambda}$ would need to remain
  inside $\beta$'s nullspace.  This would imply that

  \begin{equation*}
    VV^\dagger \bm{\lambda}^{r + 1} = \bm{\lambda}^{r + 1},
  \end{equation*}

  i.e. that $\bm{\lambda}^{r + 1}$ is an eigenvector of $VV^\dagger$
  for an infinite number of integers $r$ (the exponentiation is to be
  understood as a point wise operation).  However, since there can only
  be a finite number of (unit length) eigenvectors, this cannot be the
  case unless every eigenvalue $(\lambda_1, \ldots, \lambda_n)$ were
  equal.

  We see from this example that the collection of $\VAR(1)$ systems
  which are not persistent are pathological, in the sense that their
  system matrices have zero measure when viewed as a subset of $\R^{n^2}$.
\end{example}

\begin{lemma}
  \label{lem:time_lag_cancellation}
  Suppose $v(t)$ is a scalar process with unit variance and zero
  autocorrelation and let $\A(z), \B(z)$ be nonzero and strictly
  causal (i.e. $1 \le \tau_0(\A) < \infty$,
  $1 \le \tau_0(\B) < \infty$) linear filters.  Then,

@@

  \begin{align}
    \inner{\A(z)v(t)}{\B(z)v(t)} &= \sum_{\tau = 1}^\infty \sum_{s = 1}^\infty a(\tau)b(s)\E[v(t - s)v(t - \tau)]\\
    &= \sum_{\tau = \text{max}(\tau_0(\A), \tau_0(\B))}^{\text{min}(\tau_\infty(\A), \tau_\infty(\B))} a(\tau) b(\tau),\\
  \end{align}

@@

\begin{proposition}
  \label{prop:persistence_converse}
  Fix $i, j \in [n]$ and suppose $\exists k \in \anc{i} \cap \anc{j}$
  which confounds $i, j$.  Then, if $T_{ij}(z)$ is not causal we have
  $j \pwgc i$, and if $T_{ij}(z)$ is not anti-causal we have
  $i \pwgc j$.  Moreover, if Assumption \ref{ass:T_causality} is
  satisfied, then $j \pwgc i \iff i \pwgc j$.
  % Suppose that Assumption \ref{ass:T_causality} is satisfied.  Then if
  % there exists a $k$ which confounds $(i, j)$ we have
  % $i \pwgc j \implies j \pwgc i$.  Moreover, if $T_{ij}(z)$ is not
  % constant, then $i \pwgc j$.
\end{proposition}

% QUESTION: Does $\gcg$ need to be a strongly causal DAG for this proposition?

% \begin{proposition}
%   \label{prop:persistence_converse}
%   Suppose $\gcg$ is a strongly causal DAG and that $x(t)$ is
%   persistent, then if there exists a $k$ which confounds $(i, j)$ we
%   have $i \pwgc j \implies j \pwgc i$.
% \end{proposition}
\begin{proof}
  Recalling Theorem \ref{thm:granger_causality_equivalences}, consider
  some $\psi \in \H_{t - 1}^j$ and represent it as
  $\psi(t) = F(z)x_j(t)$ for some strictly causal filter $F(z)$.
  Then

  \begin{align*}
    &\inner{\psi(t) - \linE{\psi(t)}{\H_{t - 1}^i}}{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^i}}\\
    &\overset{(a)}{=} \inner{F(z)x_j(t)}{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^i}}\\
    &\overset{(b)}{=} \inner{F(z)\big(\A_{jj}(z)v_j(t) + \sum_{k \in \anc{j}}\A_{jk}(z)v_k(t)\big)}{(1 - H_i(z))\big(\A_{ii}(z)v_i(t) + \sum_{\ell \in \anc{i}}\A_{i\ell}(z)v_\ell(t)\big)}\\
    &\overset{(c)}{=} \sum_{k \in \anc{i}\cap\anc{j}}\inner{F(z)\A_{jk}(z)v_k(t)}{(1 - H_i(z))\A_{ik}(z)v_k(t)},
  \end{align*}

  where $(a)$ applies the orthogonality principle, $(b)$ expands with
  Equation \eqref{eqn:ancestor_expansion} with
  $H_i(z)x_i(t) = \linE{x_i(t)}{H_{t - 1}^i}$, and $(c)$ follows by
  performing cancellations of $v_k(t) \perp v_\ell(t)$ since by the
  contrapositive of Proposition \ref{prop:sc_graph_common_anc} we
  cannot have $i \in \anc{j}$ or $j \in \anc{i}$.

% , $(c)$ is obtained through cancellations from Lemma \ref{lem:vj_perp}, $(d)$ follows since $\H_i$ is self-adjoint, $(e)$ follows by definition of $\H_i$: $\exists H_i(z)$ such that $$H_i(z)x_i(t) = \H_i \sum_{\ell \in \anc{i}}\A_{i\ell}v_\ell(t),$$ $(f)$ follows by again expanding $x_i(t)$ with Equation \eqref{eqn:ancestor_expansion}, and finally $(g)$ by cancelling orthogonal components of $v(t)$.

  Through symmetric calculation, we can obtain the expression relevant
  to the determination of $i \pwgc j$ for $\phi \in \H_{t - 1}^i$ represented by the strictly causal filter $G(z): \phi(t) = G(z)x_i(t)$
  \begin{align*}
    &\inner{\phi(t) - \linE{\phi(t)}{\H_{t - 1}^j}}{x_j(t) - \linE{x_j(t)}{\H_{t - 1}^j}}\\
    &= \sum_{k \in \anc{i} \cap \anc{j}}\inner{G(z)\A_{ik}(z)v_k(t)}{(1 - H_j(z))\A_{jk}(z)v_k(t)},
  \end{align*}

  where $H_j(z)x_j(t) = \linE{x_j(t)}{\H_{t - 1}^j}.$

  We have therefore

  \begin{align}
      &(j \pwgc i): \exists F(z) \text{ s.t. } \sum_{k \in \anc{i} \cap \anc{j}}\inner{F(z)\A_{jk}(z)v_k(t)}{(1 - H_i(z))\A_{ik}(z)v_k(t)} \ne 0,\\
      &(i \pwgc j): \exists G(z) \text{ s.t. } \sum_{k \in \anc{i} \cap \anc{j}}\inner{G(z)\A_{ik}(z)v_k(t)}{(1 - H_j(z))\A_{jk}(z)v_k(t)} \ne 0.
  \end{align}

  The persistence condition, by Corollary
  \ref{cor:time_lag_cancellation}, ensures that for each
  $k \in \anc{i}\cap\anc{j}$ there is some $F(z)$ and some $G(z)$ such
  that at least one of the above terms constituting the sum over $k$
  is non-zero.  It remains to eliminate the possibility of
  cancellation in the sum.

  The adjoint of a linear filter $C(z)$ is simply $C(z^{-1})$, which
  recall is strictly anti-causal if $C(z)$ is strictly causal.  Using
  this, we can write

  \begin{align*}
    &\sum_{k \in \anc{i} \cap \anc{j}}\inner{F(z)\A_{jk}(z)v_k(t)}{(1 - H_i(z))\A_{ik}(z)v_k(t)}\\
    = &\sum_{k \in \anc{i} \cap \anc{j}}\inner{\A_{ik}(z^{-1})(1 - H_i(z^{-1}))F(z)\A_{jk}(z)v_k(t)}{v_k(t)}\\
  \end{align*}

  Moreover, it is sufficient to find some strictly causal $F(z)$ of
  the form $F(z)(1 - H_j(z))$ (abusing notation) since $1 - H_j(z)$ is
  causal.  Similarly for $G(z)$, this leads to symmetric expressions
  for $j \pwgc i$ and $i \pwgc j$ respectively:

  \begin{equation}
    \label{eqn:T_F}
    \sum_{k \in \anc{i} \cap \anc{j}}\inner{\A_{ik}(z^{-1})(1 - H_i(z^{-1}))F(z)(1 - H_j(z))\A_{jk}(z)v_k(t)}{v_k(t)},
  \end{equation}
  \begin{equation}
    \label{eqn:T_G}
    \sum_{k \in \anc{i} \cap \anc{j}}\inner{\A_{ik}(z^{-1})(1 - H_i(z^{-1}))G(z^{-1})(1 - H_j(z))\A_{jk}(z)v_k(t)}{v_k(t)}.
  \end{equation}

  Recall the filter from Assumption \ref{ass:T_causality}

  \begin{equation}
    T_{ij}(z) = \sum_{k \in \anc{i} \cap \anc{j}} \sigma_k^2\A_{ik}(z^{-1})(1 - H_i(z^{-1}))(1 - H_j(z))\A_{jk}(z).
  \end{equation}

  Since each $v_k(t)$ is uncorrelated through time,
  $\inner{T_{ij}(z)v_k(t)}{v_k(t)} = \sigma_k^2T_{ij}(0)$, and
  therefore we have $j \pwgc i$ if $T_{ij}(z)$ is \textit{not} causal
  and $i \pwgc j$ if $T_{ij}(z)$ it \textit{not} anti-causal.
  Moreover, we have $i \npwgc j$ \textit{and} $j \pwgc i$ if
  $T_{ij}(z)$ is a constant.  Therefore, under Assumption
  \ref{ass:T_causality} $j \pwgc i \implies i \pwgc j$.

  This follows since if $T_{ij}(z)$ is not causal then
  $\exists k > 0$ such that the $z^k$ coefficient of $T_{ij}(z)$ is
  non-zero, and we can choose strictly causal $F(z) = z^{-k}$ such
  that \eqref{eqn:T_F} is non-zero and therefore $j \pwgc i$.

  Similarly, if $T_{ij}(z)$ is not anti-causal, then
  $\exists k > 0$ such that the $z^{-k}$ coefficient of $T_{ij}(z)$ is
  non-zero, and we can choose strictly causal $G(z)$ so that
  $G(z^{-1}) = z^k$, and then \refeq{eqn:T_G} is non-zero and therefore
  $i \pwgc j$.
\end{proof}

% This proof is not correct -- there can be cancellation
% \begin{proof}
%   We will show that $j \pwgc i$, the other being symmetric.  First
%   note that by the contrapositive of proposition
%   \ref{prop:sc_graph_common_anc} we cannot have $i \in \anc{j}$ or
%   $j \in \anc{i}$ (else $k$ would not be a confounder) and therefore
%   \textit{every} $u \in \anc{i}\cap\anc{j}$ will be a confounder.

%   It is sufficient to show that $\exists \psi \in \H_{t - 1}^{(j)}$
%   such that

%   \begin{equation*}
%     \inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{(i)}}} \ne 0.
%   \end{equation*}

%   To this end, let $F(z)$ be an arbitrary but strictly causal linear
%   filter.  We apply equation \eqref{eqn:ancestor_expansion} to
%   $x_i(t)$ and $\psi \defeq F(z)x_j(t)$:

%   \begin{align*}
%     &\inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{(i)}}}\\
%     &\overset{(a)}{=} \inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{\A_{ii}(z)v_i(t) + \sum_{k \in \anc{i}}\A_{ik}(z)v_k(t)}\\
%     &\overset{(b)}{=} \inner{\sum_{k \in \anc{j}}\big(F(z)\A_{jk}(z)v_k(t) \\&- \linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}}\big)}{\A_{ii}(z)v_i(t) + \sum_{k \in \anc{i}}\A_{ik}(z)v_k(t)}\\
%     &\overset{(c)}{=} \sum_{k \in \anc{i}\cap\anc{j}}\Big(\inner{F(z)\A_{jk}(z)v_k(t)}{\A_{ik}(z)v_k(t)} \\&- \inner{\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}}}{\A_{ii}v_i(t)}\\
%     &- \sum_{\ell \in \anc{i}}\inner{\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}}}{\A_{i\ell}(z)v_\ell(t)}\Big)
%   \end{align*}

%   where in $(a)$ we have removed the $\linE{x_i(t)}{\H_{t - 1}^{(i)}}$
%   term via the orthogonality principle, in $(b)$ there is no
%   $F(z)\A_{jj}(z)v_j(t)$ term since due to $j \not\in \anc{i}$ it is
%   orthogonal to $\H_t^{(i)}$.  Finally, $(c)$ follows by applying
%   orthogonality properties of $v(t)$, as well as the fact that
%   $\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} = 0$ for
%   $k \not \in \anc{i}$.  Note that
%   $\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} \in \H_{t - 1}^{(i)}$
%   and therefore there is in general no cancellation in the final term
%   above for $\ell \in \anc{i}$.

%   \todo{This is clearly not immediately evident}

%   This is $0$ for every $F$ if and only if for all $F$ and
%   $\forall k \in \anc{i} \cap \anc{j}$ we have
%   $$\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} = 0$$ and
%   $$\inner{F(z)\A_{jk}(z)v_k(t)}{\A_{ik}(z)v_k(t)} = 0,$$ which by
%   Corollary \ref{cor:time_lag_cancellation} occurs if and only if
%   $\tau_0(\A_{jk}) \ge \tau_\infty(\A_{ik})$, which is impossible since by persistence
%   $\tau_0(\A_{jk}) < \infty$ and  $\tau_\infty(\A_{ik}) = \infty$.
% \end{proof}
